---
title: "Wikidata Query Service Traffic Analysis"
output:
  md_document:
    variant: markdown_github-autolink_bare_uris
fig_width: 14
fig_height: 10
fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.path = "figures/")
library(magrittr)
library(ggplot2)
```

This shows a breakdown of internal vs external WDQS usage for Phabricator task [T179850](https://phabricator.wikimedia.org/T179850).

## Per-second traffic/usage

```{r per_sec_requests, fig.cap="Per-second count of requests made to SPARQL & LDF endpoint during October 2017, broken down by internal vs external traffic."}
requests <- readr::read_csv("data/wdqs-oct17-per-sec.csv") %>%
  dplyr::rename(`99th percentile` = percentile99) %>%
  tidyr::gather("summary", "requests", -c(ts, traffic, destination)) %>%
  tidyr::spread(summary, requests, fill = 0) %>%
  tidyr::gather("per-hour summary", "requests", -c(ts, traffic, destination))

ggplot(requests, aes(x = ts, y = requests, color = `per-hour summary`)) +
  geom_line() +
  scale_color_brewer(palette = "Set1") +
  facet_grid(destination ~ traffic, scales = "free_y") +
  wmf::theme_facet(14, "Open Sans", strip.text.y = element_text(angle = 0)) +
  labs(
    x = "Date", y = "Requests per second",
    title = "WDQS per-second internal vs external SPARQL & LDF endpoint usage during Oct 2017",
    subtitle = "Internal means the IP address started with a \"10.\"; aggregates were made on an hourly basis"
  )
```

Most of the time there are fewer than 100 internal requests per second to the SPARQL endpoint. External usage of SPARQL endpoint varies considerably, with as many as 500 reqs/sec observed during one hour on October 14th.

## Daily traffic/usage

In general we get quite a bit of external traffic from bots to the SPARQL endpoint:

```{r data}
requests <- readr::read_csv("data/T179850.csv") %>%
  dplyr::arrange(date, traffic)
sparql <- readr::read_csv("data/T179850v2.csv") %>%
  dplyr::arrange(date, traffic)
```

### Requests

```{r by_traffic_agent_destination}
requests %>%
  dplyr::group_by(date, traffic, destination, agent = agent_type_plus) %>%
  dplyr::summarize(requests = sum(requests)) %>%
  ggplot(aes(x = date, y = requests)) +
  geom_area(color = "black") +
  facet_grid(destination + agent ~ traffic, scales = "free_y") +
  scale_y_continuous(labels = polloi::compress) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_fill_brewer(palette = "Set1") +
  wmf::theme_facet(14, "Open Sans", strip.text.y = element_text(angle = 0)) +
  labs(
    x = "Date", y = "Requests per day",
    title = "WDQS usage in 2017 by traffic, expanded agent type, and destination",
    subtitle = "Internal means the IP address started with a \"10.\"",
    caption = "In addition to the current \"bot\" detection, this includes requests from user agents that include a GitHub URL."
  )

requests %>%
  dplyr::group_by(date, traffic, destination, agent = agent_type) %>%
  dplyr::summarize(requests = sum(requests)) %>%
  ggplot(aes(x = date, y = requests)) +
  geom_area(color = "black") +
  facet_grid(destination + agent ~ traffic, scales = "free_y") +
  scale_y_continuous(labels = polloi::compress) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_fill_brewer(palette = "Set1") +
  wmf::theme_facet(14, "Open Sans", strip.text.y = element_text(angle = 0)) +
  labs(
    x = "Date", y = "Requests per day",
    title = "WDQS usage in 2017 by traffic, basic agent type, and destination",
    subtitle = "Internal means the IP address started with a \"10.\"",
    caption = "These requests were split according to the current \"bot\" detection."
  )

requests %>%
  dplyr::group_by(date, traffic) %>%
  dplyr::summarize(requests = sum(requests)) %>%
  ggplot(aes(x = date, y = requests)) +
  geom_area(color = "black") +
  facet_wrap(~ traffic, scales = "free_y") +
  scale_y_continuous(labels = polloi::compress) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_fill_brewer(palette = "Set1") +
  wmf::theme_facet(14, "Open Sans", strip.text.y = element_text(angle = 0)) +
  labs(
    x = "Date", y = "Requests per day",
    title = "WDQS usage by traffic",
    subtitle = "Internal means the IP address started with a \"10.\"",
    caption = "These requests were split according to the current \"bot\" detection."
  )
```

### Time to first byte (TTFB)

```{r ttfb}
ggplot(sparql, aes(x = date, y = total_ttfb)) +
  geom_bar(stat = "identity") +
  facet_grid(agent_type ~ traffic, scales = "free_y") +
  scale_y_continuous(labels = function(x) {
    return(gsub(" 0s", "", tolower(lubridate::seconds_to_period(x)), fixed = TRUE))
  }) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b '%y") +
  wmf::theme_facet(14, "Open Sans", strip.text.y = element_text(angle = 0)) +
  labs(
    x = "Date", y = "Total time to first byte (s)",
    title = "SPARQL query execution time by traffic and (basic) agent type",
    subtitle = "Internal means the IP address started with a \"10.\""
  )

ggplot(sparql, aes(x = date, y = total_ttfb)) +
  geom_bar(stat = "identity") +
  facet_grid(agent_type_plus ~ traffic, scales = "free_y") +
  scale_y_continuous(labels = function(x) {
    return(gsub(" 0s", "", tolower(lubridate::seconds_to_period(x)), fixed = TRUE))
  }) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b '%y") +
  wmf::theme_facet(14, "Open Sans", strip.text.y = element_text(angle = 0)) +
  labs(
    x = "Date", y = "Total time to first byte (s)",
    title = "SPARQL query execution time by traffic and (extended) agent type",
    subtitle = "Internal means the IP address started with a \"10.\"",
    caption = "Extended spider detection includes UAs with GitHub URLs (for example)."
  )
```
